---
title: "01 — Clinical Data Basics (CDISC & CRF)"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Learning objectives

By the end of this tutorial you can explain:

- How **CRF data** becomes **SDTM**, then **ADaM**
- What **ADSL**, **ADAE**, and **ADLB** contain
- Why subject-level (`ADSL`) and event-level (`ADAE`, `ADLB`) structures matter for MDR

## The data pipeline

Clinical trial data typically flows:

```
CRF (site entry) → SDTM (submitted tabular) → ADaM (analysis-ready) → Dashboard
```

| Layer | Example datasets | Grain | Role in MDR |
|-------|------------------|-------|-------------|
| SDTM | DM, AE, LB | Mixed | Source / submission |
| ADaM | ADSL, ADAE, ADLB | Subject or event | Analysis & review |

**ADSL** (Subject-Level Analysis Dataset): one row per subject. Holds demographics, treatment arm, population flags (`SAFFL`, `ITTFL`).

**ADAE** (Adverse Events): one row per event. Key variables: `AEDECOD` (preferred term), `AESOC` (system organ class), `TRTEMFL` (treatment-emergent flag).

**ADLB** (Laboratory): one row per lab measurement per visit. Key variables: `PARAM`, `AVAL`, `CHG`, `ANRIND`, `ABLFL`.

## Load simulated study data

```{r}
source("../R/data_prep.R")

if (!file.exists("../data/simulated/ADSL.rds")) {
  source("../R/data_generation.R")
}

data <- load_study_data("../data/simulated")
ADSL <- data$ADSL
ADAE <- data$ADAE
ADLB <- data$ADLB
```

## Explore ADSL (subject grain)

```{r}
ADSL %>%
  select(USUBJID, ARM, AGE, SEX, RACE, SAFFL) %>%
  head(10)
```

```{r}
ADSL %>%
  count(ARM, name = "n_subjects")
```

## Explore ADAE (event grain)

```{r}
ADAE %>%
  select(USUBJID, AEDECOD, AESOC, AESEV, TRTEMFL) %>%
  head(10)
```

## Join event data to subject flags

This is the pattern used in every MDR view:

```{r}
ADAE %>%
  attach_subject_flags(ADSL) %>%
  filter(TRTEMFL == "Y") %>%
  count(ARM, AEDECOD, sort = TRUE) %>%
  head(10)
```

## Explore ADLB (lab grain)

```{r}
ADLB %>%
  select(USUBJID, PARAM, AVISIT, AVAL, CHG, ANRIND, ABLFL) %>%
  head(10)
```

## CRF terminology ↔ analysis variables

| Clinical concept | CRF / SDTM | ADaM review variable |
|------------------|------------|----------------------|
| Treatment arm | ARM | `ADSL$ARM` |
| Adverse event term | AE term | `ADAE$AEDECOD` |
| Lab result | LB result | `ADLB$AVAL` |
| Safety population | inclusion flag | `ADSL$SAFFL` |
| Treatment-emergent AE | post-dose AE | `ADAE$TRTEMFL == "Y"` |

## What to say in the interview

- "I work with **ADaM analysis datasets**, not raw CRFs. `ADSL` is my anchor — I join `ADAE` and `ADLB` on `USUBJID`."
- "For MDR, I always check **population flags** like `SAFFL` before summarizing safety."
- "Treatment-emergent flags (`TRTEMFL`) separate pre-existing events from post-treatment safety signals."
- "CDISC structure lets me build **standardized templates** that scale across studies with the same variable conventions."
