---
title: "02 — Safety Signal Methods for MDR"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5, message = FALSE, warning = FALSE)
```

## Learning objectives

- Summarize **treatment-emergent adverse events (TEAEs)** by arm
- Build **SOC/PT incidence tables** for safety review
- Detect **lab shift patterns** (normal → abnormal)
- Drill from population view to **patient profile**

## Setup

```{r}
library(dplyr)
library(ggplot2)

source("../R/data_prep.R")
source("../R/theme_clinical.R")

if (!file.exists("../data/simulated/ADSL.rds")) {
  source("../R/data_generation.R")
}

data <- load_study_data("../data/simulated")
ADSL <- data$ADSL
ADAE <- data$ADAE
ADLB <- data$ADLB
```

## Method 1: TEAE counts by arm

```{r}
summarize_teae_by_arm(ADAE, ADSL)
```

## Method 2: Subject incidence (%) — primary safety signal table

This is what medical reviewers scan first: **how many subjects** experienced each event, not just event counts.

```{r}
teae_table <- teae_incidence_by_arm(ADAE, ADSL)
teae_table
```

```{r}
ggplot(
  teae_table %>% slice_head(n = 8, by = ARM),
  aes(x = reorder(AEDECOD, pct), y = pct, fill = ARM)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = clinical_arm_colors) +
  labs(
    title = "Top Treatment-Emergent Adverse Events",
    x = NULL,
    y = "Subject incidence (%)"
  ) +
  theme_clinical()
```

## Method 3: Lab shift summaries

CtQ factor: **laboratory monitoring integrity**. Shifts from normal at baseline to abnormal post-baseline warrant review.

```{r}
summarize_lab_shifts(ADLB, ADSL)
```

```{r}
shift_plot_data <- summarize_lab_shifts(ADLB, ADSL) %>%
  filter(shift_flag)

ggplot(shift_plot_data, aes(x = PARAM, y = n, fill = ARM)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = clinical_arm_colors) +
  labs(
    title = "Clinically Relevant Lab Shifts",
    subtitle = "Normal at baseline → abnormal at latest visit",
    x = NULL,
    y = "Subjects"
  ) +
  theme_clinical()
```

## Method 4: Patient profile drill-down (MDR workflow)

Population signal → individual subject investigation:

```{r}
example_patient <- ADSL$USUBJID[1]
profile <- patient_profile(example_patient, ADSL, ADAE, ADLB)

profile$demographics %>%
  select(USUBJID, ARM, AGE, SEX, RACE, SAFFL)
```

```{r}
profile$adverse_events %>%
  select(AEDECOD, AESOC, AESEV, AESTDTC, TRTEMFL)
```

```{r}
ggplot(
  profile$labs,
  aes(x = ADY, y = AVAL, color = PARAM)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = paste("Lab trends —", example_patient),
    x = "Study Day",
    y = "Lab Value"
  ) +
  theme_clinical()
```

## RBQM / CtQ mapping

| Dashboard view | CtQ factor | Why it matters |
|----------------|------------|----------------|
| TEAE incidence | Safety reporting | Detect emerging risk by arm |
| Lab shifts | Endpoint / safety integrity | Catch hepatotoxicity patterns early |
| Population filters | Eligibility / analysis population | Ensure correct denominator |
| Patient profile | Traceability | Link aggregate signal to source subject |

## What to say in the interview

- "I start with **population-level TEAE incidence**, not raw event counts, because regulators and clinicians think in subjects-at-risk."
- "Lab shift tables connect to **CtQ laboratory monitoring** — I flag normal-to-abnormal transitions for medical review."
- "MDR is iterative: filter the safety population, spot a signal, then **drill to patient level** to validate before escalation."
- "In a regulated setting, every view ties back to a documented **Medical Data Review Plan** and traceable code."
