---
title: "03 — Building the Teal MDR Dashboard"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Learning objectives

- Understand **teal** vs raw **Shiny** for clinical apps
- Configure `init()` + `cdisc_data()` with ADSL/ADAE/ADLB
- Add standard clinical modules for MDR
- Explain **filter panels**, reproducibility, and traceability

## When to use teal vs Shiny

| Approach | Best for | Trade-off |
|----------|----------|-----------|
| **teal** | Standardized clinical dashboards, CDISC data, regulated templates | Heavier dependencies, opinionated structure |
| **Shiny** | Custom UX, lightweight demos, rapid prototypes | You build filtering/reporting yourself |

For this role, teal demonstrates you know the **industry clinical stack**; the Shiny fallback shows you can deliver without it.

## Step 1: Load data and create `cdisc_data`

```{r}
library(teal)
library(teal.modules.clinical)
library(teal.slice)

ADSL <- readRDS("../data/simulated/ADSL.rds")
ADAE <- readRDS("../data/simulated/ADAE.rds")
ADLB <- readRDS("../data/simulated/ADLB.rds")

data <- cdisc_data(
  ADSL = ADSL,
  ADAE = ADAE,
  ADLB = ADLB
)
```

`cdisc_data()` registers join keys between datasets so modules can filter consistently across ADSL, ADAE, and ADLB.

## Step 2: Add clinical modules

```{r}
modules <- modules(
  tm_t_summary(
    label = "Demographics",
    dataname = "ADSL",
    arm_var = choices_selected(
      choices = variable_choices(ADSL, c("ARM", "SEX", "RACE")),
      selected = "ARM"
    ),
    summarize_vars = choices_selected(
      choices = variable_choices(ADSL, c("AGE", "SEX", "RACE")),
      selected = c("AGE", "SEX", "RACE")
    )
  ),
  tm_t_events(
    label = "Adverse Events",
    dataname = "ADAE",
    arm_var = choices_selected(
      choices = variable_choices(ADSL, c("ARM")),
      selected = "ARM"
    ),
    llt = choices_selected(
      choices = variable_choices(ADAE, c("AEDECOD")),
      selected = "AEDECOD"
    ),
    hlt = choices_selected(
      choices = variable_choices(ADAE, c("AESOC")),
      selected = "AESOC"
    )
  ),
  tm_g_lineplot(
    label = "Lab Values Over Time",
    dataname = "ADLB",
    xvar = choices_selected(
      choices = variable_choices(ADLB, c("ADY", "AVISIT")),
      selected = "ADY"
    ),
    yvar = choices_selected(
      choices = variable_choices(ADLB, c("AVAL", "CHG")),
      selected = "AVAL"
    ),
    param_var = choices_selected(
      choices = variable_choices(ADLB, c("PARAM", "PARAMCD")),
      selected = "PARAM"
    ),
    arm_var = choices_selected(
      choices = variable_choices(ADSL, c("ARM")),
      selected = "ARM"
    )
  ),
  tm_t_pp_basic_info(
    label = "Patient Profile",
    dataname = "ADSL",
    patient_col = "USUBJID",
    vars = choices_selected(
      choices = variable_choices(ADSL, c("AGE", "SEX", "RACE", "ARM", "SAFFL")),
      selected = c("AGE", "SEX", "RACE", "ARM")
    )
  )
)
```

## Step 3: Add filter panel (MDR workflow)

Filters propagate across all modules — this is the core interactive MDR pattern:

```{r}
filters <- teal_slices(
  slice_var = c(
    teal_slices$slice_var(dataname = "ADSL", varname = "ARM", title = "Treatment Arm"),
    teal_slices$slice_var(dataname = "ADSL", varname = "SEX", title = "Sex"),
    teal_slices$slice_var(dataname = "ADSL", varname = "SAFFL", title = "Safety Population")
  )
)
```

## Step 4: Initialize and run

```{r}
app <- init(
  data = data,
  title = "Study ABC123 — Medical Data Review Dashboard",
  modules = modules,
  filter = filters
)

# shiny::shinyApp(app$ui, app$server)
```

Full runnable version: `app/app.R`

Fallback (no teal): `app/app_shiny_fallback.R`

## Traceability & QC talking points

- **Git**: version-controlled templates; every change has a commit message and peer review
- **renv**: reproducible package versions via `renv.lock`
- **Fixed seed** in `R/data_generation.R`: deterministic demo data
- **teal code export**: many modules generate reproducible R code for audit trails
- **Separation of concerns**: `R/data_generation.R` → `R/data_prep.R` → `app/`

## Run the app

From project root in R:

```r
if (!file.exists("data/simulated/ADSL.rds")) {
  source("R/data_generation.R")
}

# Primary (teal)
shiny::runApp("app")

# Fallback (pure Shiny)
shiny::runApp("app/app_shiny_fallback.R")
```

## What to say in the interview

- "I use **teal** for standardized clinical modules — demographics, AE tables, lab line plots, patient profiles — all on CDISC ADaM data."
- "The **filter panel** applies one population subset across every module, which mirrors how medical reviewers work in MDR."
- "I separate **data generation, prep, and app layers** so templates are reusable and QC-able before FPI milestones."
- "If teal isn't available in an environment, I can deliver the same MDR workflow in **pure Shiny** — the architecture stays the same."
